Can government public communications elicit undue trust? Exploring the interaction between symbols and substantive information in communications

### 2018-07-29 paper2_information and symbols_02.Rmd
Add files via upload

---
title: "The effect of symbols in government communications on citizens' trust, and its interaction with substantive information"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. Those studies that have explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology I hypothesize that symbolic elements can increase citizens' trust in two ways: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by conducing citizens to pay less attention to the substantive information in the communication. As a result citizens disregard logically unpersuasive information that could decrease their trust. I examine these expectations via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was stronger among those who were exposed to the logically unpersuasive policy plans. The latter were also distracted by the symbols and thus less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their underestimation of public-service performance. One of the main remedies proposed for this is greater investment in direct government communications to citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@deacon_2001; @eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant public administration research on the effect of government public communications on citizens attitudes has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on the performance of organizations and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

@@

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information of the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions).

@@

By comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the names of US federal agencies not only represent information about their activities, but may also carry specific symbolic meaning and positive/negative "affect" that can shape citizens' support for government policies (depending on citizens' political party identification). These two studies provide examples for the role of symbols in shaping citizens' attitudes. However, they did not effectively disentangle the effects of the symbolic and informational aspects of the public communications studied.

A study by @karens_2016 did examine the distinct effect of symbols in communications on citizens' trust. The scholars surveyed university students from three European countries and examined their trust in two policy plans of the European Commission, and it's shaping by the appearance of the organization's logo in the communication of the policy plans. They found that including the Commission’s logo significantly increased citizens' trust in these policies. Their results show an impressive effect size equivalent to 0.7 standard deviations. [^footnote_karens]

[^footnote_karens]: The effect size they found was smaller with regard to the "air quality" policy compared with the "digital skills" policy which were presented (Cohen's d = 0.7 vs. 0.6). Also, the effect was smaller among subjects from the Netherlands, compared with those from Belgium and Poland (Cohen's d = 0.5 vs. 0.7).

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared citizens who were personally exposed to prolonged power outages and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the assessed government quality.

@@

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" communications [@petty_1986; @petty_2011]. In addition to its general theoretical application in psychological research, this model has been applied to studies on persuasion in various areas, including commercial advertising [@petty_1983; @swasy_1985; @dotson_2000; @drge_1989; @wheeler_2005; @moore_1987], public health campaigns [@wilson_2007; @angst_2009; @ruiter_2001; @briol_2006], media coverage [@petty_2002; @vanderwurff_2016] and media entertainment [@young_2008; @slater_2002]. Yet, thus far, the relevance of ELM has not been directly examined in relation to citizens' responses to government communications.

According to ELM, the responses of individuals to persuasive communications depends on how much effort they devote to think about the content of the message (or "the amount of thinking" according to ELM terminology). Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion via the central route is expected to be more enduring, resistant and predictive of behavior, compared with peripheral route. Of course, often people are located in between the two ends of the continuum, and thus persuasion is determined by a mixture of these processes.

[^footnote_dual_process_theories]: The conceptual distinction between the "central and peripheral routes for persuasion" largely resembles the distinctions between heuristic and systematic processing [@chaiken_1980], and systems 1 and 2 [@kahneman_2011].

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]. The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuade people directly through the peripheral route, yet at the same time they can also act as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

[^footnote_multifunction_factors]: @petty_1986 have suggested that we are more likely to find complex relations of multiple mechanisms in those situations where people are located in the middle of the elaboration continuum. When people are neither extremely motivated or unmotivated to think, they may rely on different elements in the communications to decide whether the message is worth or needs considering.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust via the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).[^footnote_interaction_strong_information]

@@

[^footnote_interaction_strong_information]: Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans. Given the focus of this study on the interaction between symbols and ill-conceived policies, the research design does not directly addresses this latter interaction.

@@

*H<sub>4</sub> - The abovementioned moderated effect of symbolic elements on citizens' trust H<sub>2</sub> is likely to be mediated by their negative effect on citizens’ amount of thinking H<sub>3</sub> (a moderated mediation hypothesis).*

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans. Such communications can be found in publicized reports, pamphlets, websites, new media pages and ads. I focus on communication of future policy plans, as opposed to communications on performance of existing policies. This allows me to utilize a real-world empirical case without compromising the reliability of the manipulation of the content. This is because most citizens do not have prior knowledge of organizations' future policy plans, and thus they will be more willing to trust a fictional communication of an organization with regard to its future policy.

@@

[^footnote_relevance_paper]: The methodological aspects which are relevant for examining the additional hypotheses (i.e. the interaction between symbols with personal relevance) are described in a separate paper. An English translation of the full survey is available in the paper's supplementary appendix.

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM) that is responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM is a good candidate for this study due to its significant investment, over the past years, in public communications. In these communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (Figure 2a). This logo consists of a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element which I focused on, is the color green, which is strongly associated with environmental protection. The EPM logo and the green color are used in almost all of EPM’s visual communications, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, who previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (Figure 2b) was launched in 2010, and included a series of advertisements in which Friedman was presented with a head covered in green grass, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (Figure 2c) was launched in 2017, and focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

In the course of the survey experiment, I was able to confirm that the EPA logo and the images of the celebrities from the campaigns were familiar to the participants. A total of `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize the EPA logo to some extent, and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`% , respectively). In robust analyses I further exclude those who do not recognize these symbols.[^footnote_prior_studies]

@@

[^footnote_prior_studies]: In two prior online survey tests, which I conducted before the experiment, I collected further information about the affect and symbolic associations of these symbols. These results are reported in the supplementary appendix and substantiate my empirical assumption that these elements evoke positive emotions in Israeli citizens, and that citizens automatically associate them with EPM and its policies. These results also indicate that the images of the celebrity comedians from EPM campaigns are perceived as humoristic, and accordingly, they are expected to evoke a positive mood in citizens.

@@

a. EPM logo.
b. Tal Friedman, comedian,  EPM campaign “starting to think green” (2010): "Isn't money growing on trees? Moving to environmental consumption and saving at least 6,000 NIS per year!", "starting to think green", "The Environmental Protection Ministry".
c. Ido Rosenblum, comedian, EPM campaign  “taking every bag seriously” (2017): "Did you for forget your reusable carrier bag at home again? A tip from Ido: Hang it on the door handle and you will not leave without it", "Taking every bag seriously", "The Environmental Protection Ministry".

@@

[^footnote_relevance_manipulation]: This section also included a manipulation of participants' perceived personal relevance. Participants in the treatment group were asked a series of questions about environmental government policy, whereas participants in the control group were asked instead equivalent questions about their field of occupation.

Per each policy, subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). Each subject, moreover, was exposed to one strong and one weak policy communication. So for example a subject who first received a strong air-pollution policy plan, later received the weak recycling policy plan, and vice versa. The actions which were mentioned in the condition of the strong policy plans were taken from the real EPM work plan and from other reports and official EPM sources. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures that are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were to decrease the supervision of polluting factories, and raise citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is included in **Appendix A**.

In addition to information manipulation, the participants were randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment ("real symbols"), control-"no symbols" and control-"fake symbols". In the symbol treatment condition, the two policies (again, one weak and one strong) were displayed with the above-mentioned three symbolic elements: the titles and the footer were colored green, the EPM logo was displayed at the top of the page, and the celebrities' images from the campaigns were attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags was attached to the recycling policy; Friedman's image from the "starting to think green" campaign was attached to the air-pollution policy. In the control no-symbols condition, the two policies were displayed in a minimal black and white design, and without the logo and the celebrities’ images. Finally, in the control fake-symbols condition, the green color was replaced with blue, the EPM logo was replaced by a fake logo and the images of celebrities from the campaigns were replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition was to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they would resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which could be attached to the government ministry and environmental issues.  In that sense, the "fake" symbols could be regarded as a placebo. The symbols conditions are displayed in **Appendix B**. Further details about the images in the real and fake symbols conditions are found in the supplementary appendix. Each subject received two policies which were presented under the same symbol condition, but under different information conditions (i.e. one strong and one weak). This allowed me later to analyze the interaction between these two manipulations both between and within-subjects.

@@

Each policy plan was followed by six questions about respondents' trust in the policy. Next, the subjects underwent a memory test on the last policy that was presented to them, and they were asked about the extent to which they elaborated on it and about the personal relevance of the policy plans. The following step included an Instructional Manipulation Check (IMC) item, intended to assess participants’ diligence. After a set of demographic questions, I conducted a manipulation check, in which I asked the participants whether they recognize the Ministry's logo, and the two celebrities’ images. Finally, the participants were debriefed about the manipulations, and the possible deception. An English translation of the full survey is available in the supplementary appendix.

Altogether, for each policy plan, the survey experiment has a factorial design of 3 (symbols: real symbols, fake-symbols, no-symbols) x 2 (information: strong vs. weak policy). The assignment to the different experimental groups is graphically presented in **Figure 3**. The experimental groups are fairly balanced with respect to demographics, and pre-manipulation variables. Throughout the rest of the paper, I will analyze the results at the policy level, in order to simplify the analyses. Accordingly, I transformed the dataset into a multilevel structure of policies nested within subjects.[^footnote_combined_analyses]

[^footnote_combined_analyses]: In the supplementary appendix I further replicate the analyses below, while adding the interactions between the manipulations and the policy plans. In these models, the interactions between the policy plans and the symbols are far from significance, with no major changes to the results.

@@

[^footnote_elaboration]: In addition, I included in the survey a battery of five items that measure participants' subjective perception of their level of elaboration. I initially sought to use these items as an additional measure for elaboration. However, I decided to forgo this subjective measure, due to its lack of correlation with the reaction time and memory test score, and its possible bias by people's disinclination to admit that they did not thoroughly read the policy plans.

>>>>

---
title: "Can government public communications elicit undue trust? Exploring the interaction between symbols and substantive information in communications"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. Those studies that have explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the social psychology Elaboration Likelihood Model I hypothesize that symbols increase citizens' trust in two ways: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by conducing citizens to pay less attention to the substantive information in the communication. As a result, citizens disregard logically unpersuasive information that could decrease their trust. I examine these expectations via a survey experiment, focusing on the Israeli Environment Protection Ministry. The results support the hypotheses and suggest that the effect of symbolic elements is stronger when communications include logically unpersuasive information. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their underestimation of public-service performance. One of the main remedies proposed for this is greater investment in direct government communications to citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant public administration research on the effect of government public communications on citizens attitudes has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on the performance of organizations and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

@@

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information in the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions).

@@

In comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public US Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the names of US federal agencies not only represent information about their activities, but may also carry specific symbolic meaning and positive/negative "affect" that can shape citizens' support for government policies (depending on citizens' political party identification). While these two studies provide examples for the role of symbols in shaping citizens' attitudes, they did not effectively disentangle the effects of the symbolic and informational aspects of the public communications studied.

A study by @karens_2016 did examine the distinct effect of symbols in communications on citizens' trust. The scholars surveyed university students from three European countries and examined their trust in two policy plans of the European Commission, and it's shaping by the appearance of the organization's logo in the communication of the policy plans. They found that including the Commission’s logo significantly increased citizens' trust in these policies. Their results show an impressive effect size equivalent to 0.7 standard deviations, on average.

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared citizens who were personally exposed to prolonged power outages and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the government quality assessed.

@@

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" communications [@petty_1986; @petty_2011]. In addition to its general theoretical application in psychological research, this model has been applied to studies on persuasion in various areas, including commercial advertising [e.g. @petty_1983; @dotson_2000; @wheeler_2005; @moore_1987], public health campaigns [@wilson_2007; @angst_2009; @ruiter_2001; @briol_2006], media coverage [@petty_2002; @vanderwurff_2016] and media entertainment [@young_2008; @slater_2002]. So far however the relevance of ELM has not been directly examined in relation to citizens' responses to government communications.

According to ELM, the responses of individuals to persuasive communications depends on how much effort they devote to thinking about the content of the message (or "the amount of thinking" according to ELM terminology). Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion via the central route is expected to be more enduring, resistant and predictive of behavior, compared with the peripheral route. Of course, often people are located in between the two ends of the continuum, and thus persuasion is determined by a combination of these processes.

[^footnote_dual_process_theories]: The conceptual distinction between the "central and peripheral routes for persuasion" largely resembles the distinctions between heuristic and systematic processing [@chaiken_1980].

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]. The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuade people directly through the peripheral route, yet at the same time they can also act as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and able to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust via the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).[^footnote_interaction_strong_information]

@@

[^footnote_interaction_strong_information]: Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans.

@@

*H<sub>4</sub> - The abovementioned moderated effect of symbolic elements on citizens' trust (H<sub>2</sub>) is likely to be mediated by their negative effect on citizens’ amount of thinking (H<sub>3</sub>) - a moderated mediation hypothesis.*

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans. Such communications can be found in publicized reports, pamphlets, websites, new media pages and ads. I focus on communication of future policy plans, as opposed to communications on performance of existing policies. This allows me to utilize a real-world empirical case, yet still to manipulate the content of the communication in a reliable manner. Most citizens do not have prior knowledge of organizations' future policy plans, and will therefore be more willing to trust a fictional communication of an organization with regard to its future policy.

@@

[^footnote_relevance_paper]: The methodological aspects which are relevant for examining the interaction between symbols with personal relevance are described in a separate paper.

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM) that is responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in the EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, the EPM is a good candidate for this study due to its significant investment, over the past years, in public communications. In these communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (Figure 2a). This logo consists of a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element which I focused on, is the color green, which is strongly associated with environmental protection. The EPM logo and the green color are used in almost all of EPM’s visual communications, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, who previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (Figure 2b) was launched in 2010, and included a series of advertisements in which Friedman was presented with a head covered in green grass, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (Figure 2c) was launched in 2017, and focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

In the course of the survey experiment, I was able to confirm that the EPA logo and the images of the celebrities from the campaigns were familiar to the participants. A total of `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize the EPA logo to some extent, and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`%, respectively). In robust analyses I further exclude those who do not recognize these symbols.[^footnote_prior_studies]

@@

[^footnote_prior_studies]: In two prior surveys, I collected further information about the affect and symbolic associations of these symbols. These results are reported in the supplementary appendix and substantiate my empirical assumption that these elements evoke positive emotions, and that citizens automatically associate them with the EPM.

@@

a. EPM logo.
b. Tal Friedman, comedian,  EPM campaign “starting to think green” (2010): "Isn't money growing on trees? Moving to environmental consumption and saving at least 6,000 NIS per year!", "starting to think green", "The Environmental Protection Ministry".
c. Ido Rosenblum, comedian, EPM campaign  “taking every bag seriously” (2017): "Did you forget your reusable carrier bag at home again? A tip from Ido: Hang it on the door handle and you will not leave without it", "Taking every bag seriously", "The Environmental Protection Ministry".

@@

[^footnote_relevance_manipulation]: This section also included a manipulation of participants' perceived personal relevance. Participants in the treatment group were asked a series of questions about environmental government policy, whereas participants in the control group were asked equivalent questions about their field of occupation.

Per each policy, subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). Each subject, moreover, was exposed to one strong and one weak policy communication. So for example a subject who first received a strong air-pollution policy plan, later received the weak recycling policy plan, and vice versa. The actions which were mentioned in the condition of the strong policy plans were taken from the real EPM work plan and from other reports and official EPM sources. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures that are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the strong air-pollution policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were to decrease the supervision of polluting factories, and raise citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is included in **Appendix A**.

In addition to information manipulation, the participants were randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment ("real symbols"), control-"no symbols" and control-"fake symbols". In the symbol treatment condition, the two policies (again, one weak and one strong) were displayed with the above-mentioned three symbolic elements: the titles and the footer were colored green, the EPM logo was displayed at the top of the page, and the celebrities' images from the campaigns were attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags was attached to the recycling policy; Friedman's image from the "starting to think green" campaign was attached to the air-pollution policy.

In the control no-symbols condition, the two policies were displayed in a minimal black and white design, and without the logo and the celebrities’ images. In the control fake-symbols condition, the green color was replaced with blue, the EPM logo was replaced by a fake logo and the images of celebrities from the campaigns were replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition was to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they would resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which could be attached to the government ministry and environmental issues.  In that sense, the "fake" symbols could be regarded as a placebo.

The symbols conditions are displayed in **Appendix B**. Further details about the images in the real and fake symbols conditions are found in the supplementary appendix. Each subject received two policies which were presented under the same symbol condition, but under different information conditions (i.e. one strong and one weak). This allowed me later to analyze the interaction between these two manipulations both between and within-subjects.

@@

Each policy plan was followed by six questions about respondents' trust in the policy. Next, the subjects underwent a memory test on the last policy that was presented to them, and they were asked about the extent to which they elaborated on it and about the personal relevance of the policy plans. The following step included an Instructional Manipulation Check (IMC) item, intended to assess participants’ diligence. After a set of demographic questions, I conducted a manipulation check, in which I asked the participants whether they recognize the Ministry's logo, and the two celebrities’ images. Finally, the participants were debriefed about the manipulations, and the possible deception. An English translation of the full survey is available in the supplementary Appendix.

Altogether, for each policy plan, the survey experiment has a factorial design of 3 (symbols: real symbols, fake-symbols, no-symbols) x 2 (information: strong vs. weak policy). The assignment to the different experimental groups is graphically presented in **Figure 3**. The experimental groups are balanced with respect to demographics, and pre-manipulation variables. Throughout the rest of the paper, I will analyze the results at the policy level, in order to simplify the analyses. Accordingly, I transformed the dataset into a multilevel structure of policies nested within subjects.[^footnote_combined_analyses]

[^footnote_combined_analyses]: In the supplementary Appendix I further replicate the analyses below, while adding the interactions between the manipulations and the policy plans. The interactions between the policy plans and the symbols are far from significance, with no major changes to the results.

@@

[^footnote_elaboration]: In addition, I included in the survey an index for participants' perceived elaboration. However, I decided to forgo this subjective measure, due to its lack of correlation with the other measures, and its possible bias by social desirability.
